Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
EAGGA

@@

\subsubsection{A-posteriori}
\label{sec-moo-post}
The disadvantage of a-priori methods is the missing knowledge of the interplay between a hyperparameter configuration and the trained model's
performance across objectives:
we can either restrict the search space to enforce some will, such as only using a maximum of 50\% of features, optimize for performance and use the
resulting model, or leave the search space unrestricted, adjust the loss function to incorporate all objectives and take its optimum.
In neither case do we know the impact our a-priori trade-off has on the final performance.
In practical applications, the impact on performance across all objectives is what matters.
This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
but keep the set of non-dominated solutions instead of just one that satisfies the trade-off criterion.
This is beneficial as it makes the relationship visible, a practicioner can now see, for instance, that a slight decrease in one objective might translate to
a significant improvement in another that would have been missed if the problem was optimized with a-priori methods.
\\
Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
\begin{enumerate}
  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augemented Tchebycheff function as
        scalarization to ensure the Pareto front is explored sufficiently.
  \item Fitting one surrogate per objective, then either using one acquisition function per surrogate to return a set of promising candidate query points, or using
        one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises the expected improvement
        of hypervolume.
\end{enumerate}

@@

%likely shouldn't be an entire section
\section{EAGGA}
- NSGA-II inspired evolutionary + genetic algorithm
- uses AUC-ROC as performance metric and 3 interpretation metrics
  * NF: rel. \# of features used in model
  * NI: rel. \# of pairwise feature interaction effects
  * NNM: rel. \# of non-monotone feature effects
- aims to find high performing models (AUC) with low NF, NI, and NNM -> ensures better interpretability than simply optimising for performance
  but creates very large objective space $\check{\boldsymbol\Lambda}$ to be optimised over with
  tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
  with
  * $\boldsymbol\lambda\in\boldsymbol\Lambda$ ``regular'' model HP config
  * $\boldsymbol{s}\in\{0,1\}^p$ vector denoting feature usage
  * $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ matrix denoting pairwise interactions
  * $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ vector denoting monotonicity constraint of each feature (-1 decreasing, 0 none, 1 increasing)
- thus would require to solve $\argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} (GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}))$
- \citep[pp. 540f]{EAGGA}
- introduces equivalence relation $R$ ``allowed to interact'' on features that significantly reduces search space for more efficient optimization
- thus: augmented search space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising model HP space and group structure space
  * each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
    - $G_1$ set of excluded features
    - $\forall k=2,...,g:G_k=(E_k,M_{E_k})$ tuple
      * $E_k$ set of features allowed to interact with each other
      * $M_{E_k}$ monotonicity constraint of entire group $k$
- EA on model hyperparameters $\boldsymbol\Lambda$
- GGA on group structure space $\mathcal{G}$ with adapted mutation and crossover operators
- \citep[pp. 541f]{EAGGA}
- futher introduces special initialisation of group structures to increase sample-efficiency of EAGGA algorithm compared to random init
- \citep[pp. 542f]{EAGGA}

>>>>

\usepackage{enumitem}  % nested enumerations, cf. https://tex.stackexchange.com/a/126756

@@

\subsubsection{A-posteriori}
\label{sec-moo-post}
The disadvantage of a-priori methods is the missing knowledge of the interplay between a hyperparameter configuration and the trained model's
performance across objectives:
we can either restrict the search space to enforce some will, such as only using a maximum of 50\% of features, optimize for performance and use the
resulting model, or leave the search space unrestricted, adjust the loss function to incorporate all objectives and take its optimum.
In neither case do we know the impact our a-priori trade-off has on the final performance.
In practical applications, the impact on performance across all objectives is what matters.
This is the main advantage of a-posteriori methods: instead of implicitly defining a trade-off prior to training, we still evaluate multiple HP configurations
but keep the set of non-dominated solutions instead of just one that satisfies the trade-off criterion.
This is beneficial as it makes the relationship visible, a practicioner can now see, for instance, that a slight decrease in one objective might translate to
a significant improvement in another that would have been missed if the problem was optimized with a-priori methods.
\\
Aside from the usual baselines grid and random search there are multi-objetive Bayesian Optimization adaptations, mainly using one of two approaches:
\begin{enumerate}[label*=\arabic*.]
  \item Fitting a single surrogate model on scalarized objectives. \citet[pp. 54-56]{ParEGO} propose \textit{ParEGO}, which employs the augemented Tchebycheff function as
        scalarization to ensure the Pareto front is explored sufficiently.
  \item Fitting one surrogate per objective, then either 
  \begin{enumerate}[label*=\arabic*.]
    \item using one acquisition function per surrogate to return a set of promising candidate query points or
    \item using one overall acquisition function to aggregate the surrogates, such as \textit{EHI} as proposed by \citet[pp. 8f]{EHI}, which maximises
          the expected improvement of hypervolume.
  \end{enumerate}
\end{enumerate}

@@

%likely shouldn't be an entire section
\section{EAGGA}
EAGGA, as proposed by \citet{EAGGA}, is an NSGA-II inspired algorithm that optimizes performance along with three interpretability objectives.
The three interpretability proxies are
\begin{itemize}
  \item NF: the relative number of features used in the model
  \item NI: the relative number of pairwise interaction effects in the model
  \item NNM: the relative number of non-monotone feature effects in the model
\end{itemize}
The authors argue that a sparse model as indicated by a low NF score helps interpretability by reducing the number of e.g. ALE plots necessary in post-hoc
analysis \citep[p. 540]{EAGGA}. Similarly, a low NI score reduces the interplay between features and thus the dimensionality of post-hoc analysis plots
and a low NNM avoids complex non-monotone feature effects, which is often desirable in practical use-cases such as loan approval, where a higher income should
translate to a higher chance of approval.
Lastly, performance is measured by the area under the receiver operating characteristic curve (AUC), which trades-off the true against the false positives rate.
\\
\citet[pp. 540f]{EAGGA} acknowledge that this set-up creates the large objective space $\check{\boldsymbol\Lambda}$ to be optimized over, containing
tuples $(\boldsymbol\lambda, \boldsymbol{s}, \boldsymbol{I}_s, \boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}})=\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}$
with
\begin{itemize}
  \item $\boldsymbol\lambda\in\boldsymbol\Lambda$ being the model's hyperparameter configuration
  \item $\boldsymbol{s}\in\{0,1\}^p$ being a binary vector denoting which features are used in the model
  \item $\boldsymbol{I}_s\in\{0,1\}^{p\times p}$ being a binary matrix denoting included pairwise interactions between features
  \item $\boldsymbol{m}_{\boldsymbol{I}_{\boldsymbol{s}}}\in\{-1,0,1\}^p$ being the vector denoting the monotonicity constraint of a feature (-1 decreasing, 0 none, 1 increasing)
\end{itemize}
Consequently, the optimization problem over this space would be
\begin{equation}
  \argmin_{\check{\boldsymbol\lambda}\in\check{\boldsymbol\Lambda}} \big(
    GE(\mathcal{I}_{\check{\boldsymbol\lambda}},\mathcal{D}),
    NF(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NI(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}}),
    NNM(\hat{f}_{\mathcal{D},\check{\boldsymbol\lambda}})
  \big)
  \label{eq-eagga-extended-space}
\end{equation}
The authors thus introduce the equivalence relation $R$ ``allowed to interact'' on the set of features $C$. They further denote all features included in the model with $C_s$.
This induces a group structure of equivalence classes, where features allowed to interact belong to the same class and share the same monotonicity attribute.
Using this group structure, they define a much simpler augmented serach space $\tilde{\boldsymbol\Lambda}=\boldsymbol\Lambda\times\mathcal{G}$ comprising of
the cartesian product of the model hyperparameter space $\boldsymbol\Lambda$ and the group structure space $\mathcal{G}$.
Each group structure $\boldsymbol{G}\in\mathcal{G}$ consists of $g$ groups:
\begin{itemize}
  \item $G_1=C\setminus C_s$ being the set of excluded features
  \item $G_k=(E_k,M_{E_k}), \forall k=2,...,g$ being tuples with
  \begin{itemize}
    \item $E_k\subseteq C_s$ being the set of features in group $k$, i.e. features allowed to interact with each other
    \item $M_{E_k}$ being the monotonicity constraint of group $k$
  \end{itemize}
\end{itemize}
To generate new offspring, they use a regular evolutionary algorithm on the model hyperparameters $\boldsymbol\Lambda$ and a grouping genetic algorithm (GGA) on
the group structure space $\mathcal{G}$, with adapted mutation and crossover operators.
The authors further introduce a special initialisation of the group structures to increase the sample-efficiency of their algorithm compared to random initialisation.
\citep[pp. 541-543]{EAGGA}
